---
title: "final project"
author: "Victoria Oldemburgo de Mello"
date: "23/11/2021"
output: pdf_document
---

# Summary
Our goal is to explore the "mtcars" dataset and understand the relationship between miles per gallon and type of transmission. In this doc, I will explore the dataset, run some relevant lm models, and try to draw some inferences based on the results of the analysis.

# Understanding the dataset
```{r, results="hide"}
head(mtcars)
?mtcars
```

# Exploring dataset
```{r, results="hide"}
plot(mtcars$mpg, mtcars$am)
```
```{r, results="hide"}
cor.test(mtcars$am, mtcars$mpg)
```
The variables we are interested in are correlated at 0.59. This is a good start.
It seems that manual  cars do more miles per gallon. Let's confirm with linear models.

# Question 1: Is an automatic or manual transmission better for MPG?
Let's confirm our intuition.
```{r, results="hide"}
fit1 <- lm(mpg ~ am, data = mtcars)
summary(fit1)
```

Manual cars do significantly more miles per gallon than automatic cars. 

But what if we control for other possible relevant variables? E.g.: horsepower, weight.

```{r, results='hide'}
fit2 <- lm(mpg ~ am + hp, data = mtcars)
summary(fit2)
# when we controled for horsepower, the coefficient changed but not significance or direction of relationship

fit3 <- lm(mpg ~ am + wt, data = mtcars)
summary(fit3)
# controlling for weight made the relationship between mpg and am non-significant!

# Maybe weight and type of transmission are correlated?
cor.test(mtcars$wt, mtcars$am)
# very correlated! at almost 0.7. Meaning that automatic cars are heavier. 
```

What if we control for all variables?
```{r, results='hide'}
fit5 <- lm(mpg ~ ., data = mtcars)
summary(fit5)
```
When we add all the variables to the model, none of them are significant. 

Let's stick to the model where we control for hp and weight. 
```{r, results='hide'}
fit6 <- lm(mpg ~ am + hp + wt, data = mtcars)
summary(fit6)
# controlling for the variables we tested makes the transmission type no longer significant. 
```

The relationship between mpg and am not controlling for anything was significant, meaning that manual cars perform better than automatic (more miles per gallon). However, when we control for weight and horsepower, the relationship is no longer significant. This is important, because it is possible that the first relationship was drawn by the fact that manual cars are lighter. 

# Question 2: Quantify the MPG difference between automatic and manual transmissions. 
```{r, results='hide'}
# model with transmission alone (holding everything else constant)
fit4 <- lm(mpg ~ am, data = mtcars)
summary(fit4)
# sum the intercept and the coefficient for manual cars
17.14+7.24

# model controlling for horsepower and weight
fit6 <- lm(mpg ~ am + hp + wt, data = mtcars)
summary(fit6)
# sum the intercept with the coefficient for manual cars
34.00+2.08

```
### WHEN NOT CONTROLLING FOR OTHER VARIABLES:
The intercept is 17.14, meaning that the average car does 17.14 miles with one gallon when it is automatic.
If we add the coefficient for the automatic car (7.24), we end up with 24.38, which is the amount of miles per gallon the average manual car does. 

Of course that doesn't allow us to infer that manual cars are more economic. As we saw before, this relationship is no longer significant once we control for weight. Thus, further tests would be needed if we wanted to make any conclusions about the relationship between transmission and gas consumption.

### WHEN CONTROLLING FOR HORSEPOWER AND WEIGHT:
The intercept is 34.00 and the coefficient of transmission (although not significant) is 2.08. This means that the average miles per gallon of an automatic car is 34.00 and the average miles per gallon for a manual car is 36.08. This is true when we control for horsepower and weight. 

# Diagnostics
Let's do residual plots for both models considered to see which one has a better fit. 
```{r}
# model with no controls
resid_simple <- resid(fit4)
# model with two controls
resid_full <- resid(fit6)

plot(mtcars$am, resid_simple) 
plot(mtcars$am, resid_full) 

```

The scales are different, but we can tell that the first model (not controlling for any other variables) has larger residuals than the "full" model - model that controls for wt and hp. 

# Conclusion
All evidence gathered up to this point suggests that a linear model that accounts for weight and horsepower are better suited for understanding the relationship between type of transmission and miles per gallon performance. In all models, the direction of the relationship was such that manual cars performed better (more miles per gallon) than automatic cars. However, this relationship was only significant in the model that did not account for the control variables. 






